---
title: "Wildlife Crime Dashboard"
author: "Alex Ley"
format: 
  dashboard:
    theme: journal
    orientation: columns
editor: visual
embed-resources: true
---

```{r setup}
#| message: false

library(tidyverse)
library(ggplot2)
library(knitr)
library(dplyr)
library(sf)
library(rnaturalearth)
library(lubridate)
library(DT)

setwd("/Users/alexley/Documents/STAT408/Semester Project")

incidents = read_csv("Data/incident-data-3678.csv")
incidents = incidents[, 1:16]
incidents = incidents |>
  select(-"Primary Source", -"Source Type", -"Additional Sources", -"Method of Concealment", -"Name of Organisation Providing Information", -"Methods - Details")

names(incidents) = c("ID", "incident_type", "country", "date", "subject", "transport_mode", "where_found", "detection", "outcome", "arrests")

incidents = incidents |>
    mutate(incident_type = str_remove(incident_type, "^[0-9]+\\.\\s*"))
```

# Overview

## Column {width=55%}
### Row 

The illegal wildlife trade (IWT) is considered to be one of the largest illegal markets in the world, drawing the attention of numerous organized crime groups (Martínez and Alonso, 2021). This market is estimated to be worth 7-23 billion USD per year (Nellemann et al., 2016). Countries around the world experience wildlife crime with many government agencies working to try to prevent it from occurring. Despite this effort, many species are still targeted for various parts, including rhino ivory, tiger bone marrow, bushmeat, and more. This project helps us visualize the number of wildlife trade incidents that are occuring, countries they are most common in, and arrests being made.

### Row 
```{r}
# Trends over time

incidents <- incidents |>
  mutate(date = as.Date(date))

trends_year <- incidents |>
  mutate(year = lubridate::year(date)) |>
  count(year, name = "n") |>
  arrange(year)

ggplot(trends_year, aes(x = year, y = n)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = unique(trends_year$year)) +
  labs(
    x = "Year",
    y = "Number of Seizures",
    title = "Wildlife Crime Seizure Trends Over Time"
  ) +
  theme_minimal()
```

## Column {width=45%}
### Row {height=35%}
```{r}
#| message: false
# Value boxes (summary stats)
total_seizures <- format(nrow(incidents))

total_arrests = sum(incidents$arrests, na.rm = TRUE)
```

```{r}
#| content: valuebox
#| title: "Total Seizures"

list(
  icon = "exclamation-triangle",
  color = "danger",
  value = total_seizures
)
```

```{r}
#| content: valuebox
#| title: "Total Arrests"

list(
  icon = "person-badge",
  color = "warning",
  value = total_arrests
)
```

### Row {height=65%}
```{r}
# Pie Chart Most common seizure type
seizure_type = incidents |>
  count(incident_type, name = "n") |>
  arrange(desc(n))

ggplot(seizure_type, aes(x = "", y = n, fill = incident_type)) +
  geom_col(width = 1) +
  coord_polar("y") +
  theme_void() +
  labs(
    title = "Distribution of Seizure Types",
    fill = "Incident Type"
  ) +
  scale_fill_brewer(palette = "Spectral")
```


# Global Patterns

## Column {width=55%}
### Row {height=45%}

**DATA SUMMARY**

These figures depict the countries where most incidents occur and the distribution of the various types of incidents. Asian countries often hold the highest demand for illegal wildlife products and also homes a lot of high demand species. Most incident types are classified as a seizure and the second most common incident type is poaching / illegal harvesting.


```{r}
#Top 10 Countries
counts2 = incidents |> count(country)
top_counts = counts2 |>
  slice_max(n, n = 10)   # top 10 countries

ggplot(top_counts, aes(x = reorder(country, n), y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    x = "Country", y = "Count",
    title = "Wildlife Crime Incidents (Top 10 Countries)"
  )
```


### Row {height=55%}
```{r}
# Map of Incidents by Country

# Count number of seizures per country
seizures_country = incidents |>
  count(country, name = "num_seizures")

# Load world shapefile
world = ne_countries(scale = "medium", returnclass = "sf")

# Join seizure data to world map
world_seizures = world |>
  left_join(seizures_country, by = c("name" = "country"))

# Replace NA values with 0 (countries with no seizures recorded)
world_seizures$num_seizures[is.na(world_seizures$num_seizures)] = 0

# Plot map
ggplot(world_seizures) +
  geom_sf(aes(fill = num_seizures), color = "gray70", size = 0.1) +
  scale_fill_gradient( 
    low = "lightyellow", 
    high = "darkred", 
    name = "Seizures"
  )+
  coord_sf(expand = FALSE) +
  labs(
    title = "Wildlife Crime Seizures by Country",
    subtitle = "Number of recorded incidents",
    caption = "Source: Wildlife Trade Portal"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank()
  )
```


## Column {width=45%}
```{r}
#| fig-width: 14
#| fig-height: 18
# Top 50 Countries by Incident Type
top50 = incidents |>
  count(country, sort = TRUE) |>
  slice_head(n = 50)

incidents_top50 = incidents |>
  filter(country %in% top50$country)

ggplot(incidents_top50, aes(x = country, fill = incident_type)) +
  geom_bar() +
  coord_flip() +
  labs(
    x = "Country",
    y = "Count",
    title = "Type of Incident by Country"
  ) +
  scale_fill_brewer(palette = "Spectral")
```



# Transport Methods and Arrests

## Column
### Row
**DATA SUMMARY**

Most incidents did not report the type of transport used for smuggling, however, the most common was a vehicle on land, air, or sea. The number of arrests is related to the number of incidents in each country, with India in the lead, and then a large decrease of arrests following. This figure portrays the importance of increasing law enforcement and limiting current corruption as arrests are relatively low compared to the number of seizures occuring. Many arrests lead to confiscation, imprisonment, or fines, all of which tend to be relatively inconsequential. 

### Row
```{r}
# Methods of Transport
transport_methods = incidents |>
  mutate(transport_mode = ifelse(is.na(transport_mode),
                                 "Unknown",
                                 transport_mode)) |>
  count(transport_mode, name = "num_transport")

ggplot(transport_methods, aes(x = reorder(transport_mode, num_transport), y = num_transport)) +
  geom_bar(stat = "identity") +
  labs(x = "Type of Transport", y = "Count",
       title = "Type of Transport Used for Smuggling")
```

## Column
### Row
```{r}
# Arrests per Country

arrests_by_country = incidents |>
  group_by(country) |>
  summarise(total_arrests = sum(arrests, na.rm = TRUE)) |>
  filter(total_arrests > 100)

ggplot(arrests_by_country, 
       aes(x = reorder(country, total_arrests), y = total_arrests)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Country", y = "Number of Arrests",
       title = "Total Arrests by Country")
```

### Row
```{r}
# Outcome of Arrests 

arrest_outcome = incidents |>
  mutate(
    outcome = case_when(
      outcome == "-" ~ "No details given",
      str_detect(outcome, "Abscond") ~ "Absconded",
      str_detect(outcome, "Charged - no further details") ~ "Charged - No further details",
      str_detect(outcome, "Confiscation & arrest") ~ "Confiscation & Arrest",
      str_detect(outcome, "Conviction - court fine or imprisonment") ~ "Conviction - court fine & imprisonment",
      str_detect(outcome, "Conviction - Imprisonment") ~ "Conviction - imprisonment",
      str_detect(outcome, "Conviction - suspension") ~ "Conviction",
      str_detect(outcome, "Intelligence generated") ~ "Case discontinued",
      str_detect(outcome, "Information compromised") ~ "Case discontinued",
      TRUE ~ outcome
    )
  ) |>
  count(outcome, name = "num_outcome")

ggplot(arrest_outcome, aes(x = outcome, y = num_outcome))+
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Type of Outcome", y = "Count",
       title = "Distribution of Different Arrest Outcomes")
```


# Data

## Row
Data is derived from Wildlife Trade Portal and covers the years 2015-2025. This table only shows the first 500 out of 20,587 entries.

```{r}
incidents_clean <- incidents |>
  rename(
    ID = ID,
    `Incident Type` = incident_type,
    Country = country,
    Date = date,
    Subject = subject,
    `Transport Mode` = transport_mode,
    `Where Found` = where_found,
    `Detection Method` = detection,
    Outcome = outcome,
    Arrests = arrests
  )

incidents_clean |>
  arrange(ID) |>
  head(500) |>
  datatable()
```


