---
title: "Wildlife Crime Dashboard"
author: "Alex Ley"
format: 
  dashboard:
    theme: cosmo
    orientation: rows
editor: visual
embed-resources: true
---

```{r setup}
#| message: false

library(tidyverse)
library(ggplot2)
library(knitr)
library(dplyr)
library(sf)
library(rnaturalearth)
library(lubridate)
library(DT)

setwd("/Users/alexley/Documents/STAT408/Semester Project")

incidents = read_csv("Data/incident-data-3678.csv")
incidents = incidents[, 1:16]
incidents = incidents |>
  select(-"Primary Source", -"Source Type", -"Additional Sources", -"Method of Concealment", -"Name of Organisation Providing Information", -"Methods - Details")

names(incidents) = c("ID", "incident_type", "country", "date", "subject", "transport_mode", "where_found", "detection", "outcome", "arrests")

incidents = incidents |>
    mutate(incident_type = str_remove(incident_type, "^[0-9]+\\.\\s*"))
```

# Overview

## Row {height=30%}
### Column {width=60%}

The illegal wildlife trade (IWT) is considered to be one of the largest illegal markets in the world, drawing the attention of numerous organized crime groups (Martínez and Alonso, 2021). This market is estimated to be worth 7-23 billion USD per year (Nellemann et al., 2016). Countries around the world experience wildlife crime with many government agencies working to try to prevent it from occurring. Despite this effort, many species are still targeted for various parts, including rhino ivory, tiger bone marrow, bushmeat, and more. This project helps us visualize the number of wildlife trade incidents that are occuring, countries they are most common in, and arrests being made.

### Column {width=20%}
```{r}
#| message: false
# Value boxes (summary stats)
total_seizures <- format(nrow(incidents))

total_arrests = sum(incidents$arrests, na.rm = TRUE)
```

```{r}
#| content: valuebox
#| title: "Total Seizures"

list(
  icon = "exclamation-triangle",
  color = "danger",
  value = total_seizures
)
```

### Column {width=20%}
```{r}
#| content: valuebox
#| title: "Total Arrests"

list(
  icon = "person-badge",
  color = "warning",
  value = total_arrests
)
```

## Row {height=70%}
### Column {width=60%}
```{r}
# Trends over time

incidents <- incidents |>
  mutate(date = as.Date(date))

trends_year <- incidents |>
  mutate(year = lubridate::year(date)) |>
  count(year, name = "n") |>
  arrange(year)

ggplot(trends_year, aes(x = year, y = n)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = unique(trends_year$year)) +
  labs(
    x = "Year",
    y = "Number of Seizures",
    title = "Wildlife Crime Seizure Trends Over Time"
  ) +
  theme_minimal()
```

### Column {width=40%}
```{r}
# Pie Chart Most common seizure type
seizure_type = incidents |>
  count(incident_type, name = "n") |>
  arrange(desc(n))

ggplot(seizure_type, aes(x = "", y = n, fill = incident_type)) +
  geom_col(width = 1) +
  coord_polar("y") +
  theme_void() +
  labs(
    title = "Distribution of Seizure Types",
    fill = "Incident Type"
  ) +
  scale_fill_brewer(palette = "Spectral")
```


# Global Patterns

## Row
### Column 
#### Left Side Dashboard
```{r}
#| fig-width: 12
#| fig-height: 16
# Map of Incidents by Country

# Count number of seizures per country
seizures_country = incidents |>
  count(country, name = "num_seizures")

# Load world shapefile
world = ne_countries(scale = "medium", returnclass = "sf")

# Join seizure data to world map
world_seizures = world |>
  left_join(seizures_country, by = c("name" = "country"))

# Replace NA values with 0 (countries with no seizures recorded)
world_seizures$num_seizures[is.na(world_seizures$num_seizures)] = 0

# Plot map
ggplot(world_seizures) +
  geom_sf(aes(fill = num_seizures), color = "gray70", size = 0.1) +
  scale_fill_gradient( 
    low = "lightyellow", 
    high = "darkred", 
    name = "Seizures"
  )+
  coord_sf(expand = FALSE) +
  labs(
    title = "Wildlife Crime Seizures by Country",
    subtitle = "Number of recorded incidents",
    caption = "Source: Wildlife Trade Portal"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank()
  )
```

```{r}
#Top 10 Countries
counts2 = incidents |> count(country)
top_counts = counts2 |>
  slice_max(n, n = 10)   # top 10 countries

ggplot(top_counts, aes(x = reorder(country, n), y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    x = "Country", y = "Count",
    title = "Wildlife Crime Incidents (Top 10 Countries)"
  )
```


### Column 
```{r}
#| fig-width: 14
#| fig-height: 18
# Top 50 Countries by Incident Type
top50 = incidents |>
  count(country, sort = TRUE) |>
  slice_head(n = 50)

incidents_top50 = incidents |>
  filter(country %in% top50$country)

ggplot(incidents_top50, aes(x = country, fill = incident_type)) +
  geom_bar() +
  coord_flip() +
  labs(
    x = "Country",
    y = "Count",
    title = "Type of Incident by Country"
  ) +
  scale_fill_brewer(palette = "Spectral")
```



# Transport Methods and Arrests

```{r}
# Methods of Transport
transport_methods = incidents |>
  mutate(transport_mode = ifelse(is.na(transport_mode),
                                 "Unknown",
                                 transport_mode)) |>
  count(transport_mode, name = "num_transport")

ggplot(transport_methods, aes(x = reorder(transport_mode, num_transport), y = num_transport)) +
  geom_bar(stat = "identity") +
  labs(x = "Type of Transport", y = "Count",
       title = "Type of Transport Used for Smuggling")
```

```{r}
# Arrests per Country

arrests_by_country = incidents |>
  group_by(country) |>
  summarise(total_arrests = sum(arrests, na.rm = TRUE)) |>
  filter(total_arrests > 10)


ggplot(arrests_by_country, 
       aes(x = reorder(country, total_arrests), y = total_arrests)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Country", y = "Number of Arrests",
       title = "Total Arrests by Country")
```


# Data

Data is derived from Wildlife Trade Portal and covers the years 2015-2025. This table only shows the first 500 out of 20,587 entries.

```{r}
incidents_clean <- incidents |>
  rename(
    ID = ID,
    `Incident Type` = incident_type,
    Country = country,
    Date = date,
    Subject = subject,
    `Transport Mode` = transport_mode,
    `Where Found` = where_found,
    `Detection Method` = detection,
    Outcome = outcome,
    Arrests = arrests
  )

incidents_clean |>
  arrange(ID) |>
  head(500) |>
  datatable()
```


